Live Functional Programming with Typed Holes

### 2018-10-31 related-work.tex
fixed incorrect characterization of ref(tau) + invariance

The semantics borrows machinery related to type holes from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}, as discussed
at length in Sec.~\ref{sec:calculus}. The main innovation relative to this
prior work is the treatment of cast failures like holes, rather than errors.
Many of the methods developed to make gradual type systems more expressive and practical are directly relevant to future directions for \Hazel and other implementations of the ideas herein \cite{takikawa_et_al:LIPIcs:2015:5215}. For example, there has been substantial work on the problem of implementing casts efficiently \cite{herman2010space,siek2010threesomes,garcia2013threesomes}. There has also been substantial work on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,DBLP:journals/pacmpl/DevriesePP18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}. Another direction that we leave to future work is the integration of these ideas into dependently typed proof assistants, 
where evaluation is primarily at the service of equational reasoning \cite{DBLP:journals/corr/abs-1009-2789}. Several modern proof assistants, e.g. Agda \cite{norell2009dependently,norell:thesis}, Idris \cite{brady2013idris} and Beluga \cite{DBLP:conf/flops/Pientka10} already provide syntactic and static support for holes (see Sec.~\ref{sec:intro}) and interactive, hole-driven development is standard practice.

%% Although our expectation is that \Hazel will remain a pure functional language \emph{a la} \Elm~or Haskell,
%
Another interesting future direction would be to move beyond
pure functional programming and carefully integrate imperative features, e.g. ML-style references.
%
%% These require particular care, as recognized by \citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15}, but the techniques are known and we see no reason why the type safety properties  established in this paper would not be conserved, with suitable modifications to account for a store.
%
\citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15} show how to incorporate such features into gradual type theory (by treating $\mathtt{ref}(\tau)$ as invariant with respect to consistency); we expect that this approach would also work for the semantics of Sec.~\ref{sec:calculus}, i.e. it would conserve the type safety properties established there, with suitable modifications to account for a store. However, the commutativity property we establish in Sec.~\ref{sec:resumption} will not hold for a language that supports non-commutative effects. We leave to future work the task of defining more restricted special cases of the fill-and-resume operation that respects a suitable commutativity property in effectful settings (perhaps by checkpointing).

@@

We did not give type holes unique names nor did we define a commutative type hole filling operation, but we conjecture that it is possible to do so by converting casts that go from the filled hole to an inconsistent type into non-empty expression holes. We leave the details to future work.

@@

>>>>

The semantics borrows machinery related to type holes from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}, as discussed
at length in Sec.~\ref{sec:calculus}. The main innovation relative to this
prior work is the treatment of cast failures like holes, rather than errors.
Many of the methods developed to make gradual type systems more expressive and practical are directly relevant to future directions for \Hazel and other implementations of the ideas herein \cite{takikawa_et_al:LIPIcs:2015:5215}. For example, there has been substantial work on the problem of implementing casts efficiently \cite{herman2010space,siek2010threesomes,garcia2013threesomes}. There has also been substantial work on integrating gradual typing with polymorphism \cite{DBLP:conf/esop/XieBO18,DBLP:journals/pacmpl/DevriesePP18,Igarashi:2017:PGT:3136534.3110284}, and with refinement types \cite{DBLP:conf/popl/LehmannT17}.

%% Although our expectation is that \Hazel will remain a pure functional language \emph{a la} \Elm~or Haskell,
%
Another interesting future direction would be to move beyond
pure functional programming and carefully integrate imperative features, e.g. ML-style references.
%
%% These require particular care, as recognized by \citet{Siek06a} and \citet{DBLP:conf/esop/SiekVCTG15}, but the techniques are known and we see no reason why the type safety properties  established in this paper would not be conserved, with suitable modifications to account for a store.
%
\citet{DBLP:conf/esop/SiekVCTG15} show how to incorporate such features into gradual type theory; we expect that this approach would also work for the semantics of Sec.~\ref{sec:calculus}, i.e. it would conserve the type safety properties established there, with suitable modifications to account for a store. However, the commutativity property we establish in Sec.~\ref{sec:resumption} will not hold for a language that supports non-commutative effects. We leave to future work the task of defining more restricted special cases of the fill-and-resume operation that respects a suitable commutativity property in effectful settings (perhaps by checkpointing).

@@

We did not give type holes unique names nor did we define a commutative type hole filling operation, but we conjecture that it is possible to do so by replacing casts that go from the filled hole to an inconsistent type with non-empty expression holes. We leave the details to future work.

@@

\todo{integrate me}Another direction that we leave to future work is the integration of type holes into dependently typed proof assistants, 
where evaluation is primarily at the service of equational reasoning \cite{DBLP:journals/corr/abs-1009-2789}. Several modern proof assistants, e.g. Agda \cite{norell2009dependently,norell:thesis}, Idris \cite{brady2013idris} and Beluga \cite{DBLP:conf/flops/Pientka10} already provide syntactic and static support for holes (see Sec.~\ref{sec:intro}) and interactive, expression-hole-driven development is standard practice, but type holes are less well-supported.
